\pkg{cv}: An \proglang{R} Package for Cross-Validation of Regression Models

### 2024-01-02 scratch/notes_john/JSS-article/JSS-article.tex
use JSS markup \pkg() etc.

\author{
John Fox~\orcidlink{0000-0002-1196-8012}\\McMaster
University \And Georges Monette~\orcidlink{0000-0000-0000-0000}\\York
University
}
\title{\pkg{cv}: An R Package for Cross-Validation of Regression Models}

@@

The data for this example are drawn from the \textbf{ISLR2} package for
R, associated with \citet{JamesEtAl:2021}. The presentation here is
close (though not identical) to that in the original source \citep[
Secs. 5.1, 5.3]{JamesEtAl:2021}, and it demonstrates the use of the
\texttt{cv()} function in the \textbf{cv} package.\footnote{\citet{JamesEtAl:2021}
  use the \texttt{cv.glm()} function in the \textbf{boot} package
  \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
  \texttt{cv.glm()} is an independent function and not a method of a
  \texttt{cv()} generic function.}

@@

The code for this graph uses the \texttt{mse()} function from the
\textbf{cv} package to compute the MSE for each fit.

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 534539
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.22944
bias-adjusted cross-validation criterion = 19.21651
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

The \texttt{"lm"} method by default uses \texttt{mse()} as the CV
criterion and the Woodbury matrix identity to update the regression with
each fold deleted without having literally to refit the model.
Computational details are discussed in the final section of this
vignette. The function reports the CV estimate of MSE, a biased-adjusted
estimate of the MSE (the bias adjustment is explained in the final
section), and the MSE is also computed for the original, full-sample
regression. Because the division of the data into 10 folds is random,
\texttt{cv()} explicitly (randomly) generates and saves a seed for R's
pseudo-random number generator, to make the results replicable. The user
can also specify the seed directly via the \texttt{seed} argument to
\texttt{cv()}.

To perform LOO CV, we can set the \texttt{k} argument to \texttt{cv()}
to the number of cases in the data, here \texttt{k=392}, or, more
conveniently, to \texttt{k="loo"} or \texttt{k="n"}:

@@

For LOO CV of a linear model, \texttt{cv()} by default uses the
hatvalues from the model fit to the full data for the LOO updates, and
reports only the CV estimate of MSE. Alternative methods are to use the
Woodbury matrix identity or the ``naive'' approach of literally
refitting the model with each case omitted. All three methods produce
exact results for a linear model (within the precision of floating-point
computations):

@@

The \texttt{"naive"} and \texttt{"Woodbury"} methods also return the
bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an
issue for LOO CV.

This is a small regression problem and all three computational
approaches are essentially instantaneous, but it is still of interest to
investigate their relative speed. In this comparison, we include the
\texttt{cv.glm()} function from the \textbf{boot} package, which takes
the naive approach, and for which we have to fit the linear model as an
equivalent Gaussian GLM. We use the \texttt{microbenchmark()} function
from the package of the same name for the timings \citep{Mersmann:2023}:

@@

On our computer, using the hatvalues is about an order of magnitude
faster than employing Woodbury matrix updates, and more than two orders
of magnitude faster than refitting the model.\footnote{Out of
  impatience, we asked \texttt{microbenchmark()} to execute each command
  only 10 times rather than the default 100. With the exception of the
  last columns, the output is self-explanatory. The last column shows
  which methods have average timings that are statistically
  distinguishable. Because of the small number of repetitions (i.e.,
  10), the \texttt{"hatvalues"} and \texttt{"Woodbury"} methods aren't
  distinguishable, but the difference between these methods persists
  when we perform more repetitions---we invite the reader to redo this
  computation with the default \texttt{times=100} repetitions.}

@@

The \texttt{cv()} function also has a method that can be applied to a
list of regression models for the same data, composed using the
\texttt{models()} function. For \(k\)-fold CV, the same folds are used
for the competing models, which reduces random error in their
comparison. This result can also be obtained by specifying a common seed
for R's random-number generator while applying \texttt{cv()} separately
to each model, but employing a list of models is more convenient for
both \(k\)-fold and LOO CV (where there is no random component to the
composition of the \(n\) folds).

We illustrate with the polynomial regression models of varying degree
for the \texttt{Auto} data (discussed previously), beginning by fitting
and saving the 10 models:

@@

We then apply \texttt{cv()} to the list of 10 models (the \texttt{data}
argument is required):

@@

Because we didn't supply names for the models in the calls to the
\texttt{models()} function, the names \texttt{model.1},
\texttt{model.2}, etc., are generated by the function.

@@

Alternatively, we can use the \texttt{plot()} method for
\texttt{"cvModList"} objects to compare the models, though with separate
graphs for 10-fold and LOO CV:

@@

The \texttt{Mroz} data set from the \textbf{carData} package
\citep[associated with][]{FoxWeisberg:2019} has been used by several
authors to illustrate binary logistic regression; see, in particular
\citet{FoxWeisberg:2019}. The data were originally drawn from the U.S.
Panel Study of Income Dynamics and pertain to married women. Here are a
few cases in the data set:

@@

The response variable in the logistic regression is \texttt{lfp},
labor-force participation, a factor coded \texttt{"yes"} or
\texttt{"no"}. The remaining variables are predictors:

\begin{itemize}
\tightlist
\item
  \texttt{k5}, number of children 5 years old of younger in the woman's
  household;
\item
  \texttt{k618}, number of children between 6 and 18 years old;
\item
  \texttt{age}, in years;
\item
  \texttt{wc}, wife's college attendance, \texttt{"yes"} or
  \texttt{"no"};
\item
  \texttt{hc}, husband's college attendance;
\item
  \texttt{lwg}, the woman's log wage rate if she is employed, or her
  \emph{imputed} wage rate, if she is not \citep[a variable that][ show
  is problematically defined]{FoxWeisberg:2019}; and
\item
  \texttt{inc}, family income, in \$1000s, exclusive of wife's income.
\end{itemize}

We use the \texttt{glm()} function to fit a binary logistic regression
to the \texttt{Mroz} data:

@@

In addition to the usually summary output for a GLM, we show the result
of applying the \texttt{BayesRule()} function from the \textbf{cv}
package to predictions derived from the fitted model. Bayes rule, which
predicts a ``success'' in a binary regression model when the fitted
probability of success {[}i.e., \(\phi = \Pr(y = 1)\){]} is
\(\widehat{\phi} \ge .5\) and a ``failure'' if
\(\widehat{\phi} < .5\).\footnote{\texttt{BayesRule()} does some error
  checking; \texttt{BayesRule2()} is similar, but omits the error
  checking, and so can be faster for large problems.} The first argument
to \texttt{BayesRule()} is the binary \{0, 1\} response, and the second
argument is the predicted probability of success. \texttt{BayesRule()}
returns the proportion of predictions that are \emph{in error}, as
appropriate for a ``cost'' function.

The value returned by \texttt{BayesRule()} is associated with an
``attribute'' named \texttt{"casewise\ loss"} and set to
\texttt{"y\ !=\ round(yhat)"}, signifying that the Bayes rule CV
criterion is computed as the mean of casewise values, here 0 if the
prediction for a case matches the observed value and 1 if it does not
(signifying a prediction error). The \texttt{mse()} function for numeric
responses is also calculated as a casewise average. Some other criteria,
such as the median absolute error, computed by the \texttt{medAbsErr()}
function in the \textbf{cv} package, aren't averages of casewise
components. The distinction is important because, to our knowledge, the
statistical theory of cross-validation, for example, in
\citet{DavisonHinkley:1997}, \citet{BatesHastieTibshirani:2023}, and
\citet{ArlotCelisse:2010}, is developed for CV criteria like MSE that
are means of casewise components. As a consequence, we limit computation
of bias adjustment and confidence intervals (see below) to criteria that
are casewise averages.

@@

The \texttt{"glm"} method for \texttt{cv()} is largely similar to the
\texttt{"lm"} method, although the default algorithm, selected
explicitly by \texttt{method="exact"}, refits the model with each fold
removed (and is thus equivalent to \texttt{method="naive"} for
\texttt{"lm"} models). For generalized linear models,
\texttt{method="Woodbury"} or (for LOO CV) \texttt{method="hatvalues"}
provide approximate results (see the last section of the vignette for
details):

@@

To ensure that the two methods use the same 10 folds, we specify the
seed for R's random-number generator explicitly; here, and as is common
in our experience, the \texttt{"exact"} and \texttt{"Woodbury"}
algorithms produce nearly identical results. The CV estimates of
prediction error are slightly higher than the estimate based on all of
the cases.

The printed output includes a 95\% confidence interval for the
bias-adjusted Bayes rule CV criterion.
\citet{BatesHastieTibshirani:2023} show that these confidence intervals
are unreliable for models fit to small samples, and by default
\texttt{cv()} computes them only when the sample size is 400 or larger
and when the CV criterion employed is an average of casewise components,
as is the case for Bayes rule. See the final section of the vignette for
details of the computation of confidence intervals for bias-adjusted CV
criteria.

@@

As for linear models, we report some timings for the various
\texttt{cv()} methods of computation in LOO CV as well as for the
\texttt{cv.glm()} function from the \textbf{boot} package (which,
recall, refits the model with each case removed, and thus is comparable
to \texttt{cv()} with \texttt{method="exact"}):

@@

In contrast, mixed-effects models are fit to \emph{dependent} data, in
which cases as clustered, such as hierarchical data, where the clusters
comprise higher-level units (e.g., students clustered in schools), or
longitudinal data, where the clusters are individuals and the cases
repeated observations on the individuals over time.\footnote{There are,
  however, more complex situations that give rise to so-called
  \emph{crossed} (rather than \emph{nested}) random effects. For
  example, consider students within classes within schools. In primary
  schools, students typically are in a single class, and so classes are
  nested within schools. In secondary schools, however, students
  typically take several classes and students who are together in a
  particular class may not be together in other classes; consequently,
  random effects based on classes within schools are crossed. The
  \texttt{lmer()} function in the \textbf{lme4} package is capable of
  modeling both nested and crossed random effects, and the \texttt{cv()}
  methods for mixed models in the \textbf{cv} package pertain to both
  nested and crossed random effects. We present an example of the latter
  later in the vignette.}

@@

The HSB data are included in the \texttt{MathAchieve} and
\texttt{MathAchSchool} data sets in the \textbf{nlme} package
\citep{PinheiroBates:2000}. \texttt{MathAchieve} includes
individual-level data on 7185 students in 160 high schools, and
\texttt{MathAchSchool} includes school-level data:

@@

We'll use only the \texttt{School}, \texttt{SES} (students'
socioeconomic status), and \texttt{MathAch} (their score on a
standardized math-achievement test) variables in the
\texttt{MathAchieve} data set, and \texttt{Sector} (\texttt{"Catholic"}
or \texttt{"Public"}) in the \texttt{MathAchSchool} data set.

Some data-management is required before fitting a mixed-effects model to
the HSB data, for which we use the \textbf{dplyr} package
\citep{WickhamEtAl:2023}:

@@

In the process, we created two new school-level variables:
\texttt{meanses}, which is the average SES for students in each school;
and \texttt{cses}, which is school-average SES centered at its mean. For
details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.

Still following Fox and Weisberg, we proceed to use the \texttt{lmer()}
function in the \textbf{lme4} package \citep{BatesEtAl:2015} to fit a
mixed model for math achievement to the HSB data:

@@

For cluster-level CV, the \texttt{clusterVariables} argument tells
\texttt{cv()} how the clusters are defined. Were there more than one
clustering variable, say classes within schools, these would be provided
as a character vector of variable names:
\texttt{clusterVariables\ =\ c("school",\ "class")}. For cluster-level
CV, the default is \texttt{k\ =\ "loo"}, that is, leave one cluster out
at a time; we instead specify \texttt{k\ =\ 10} folds of clusters, each
fold therefore comprising \(160/10 = 16\) schools.

If the \texttt{clusterVariables} argument is omitted, then case-level CV
is employed, with \texttt{k\ =\ 10} folds as the default, here each with
\(7185/10 \approx 719\) students. Notice that one of the 10 models refit
with a fold removed failed to converge. Convergence problems are common
in mixed-effects modeling. The apparent issue here is that an estimated
variance component is close to or equal to 0, which is at a boundary of
the parameter space. That shouldn't disqualify the fitted model for the
kind of prediction required for cross-validation.

There is also a \texttt{cv()} method for linear mixed models fit by the
\texttt{lme()} function in the \textbf{nlme} package, and the arguments
for \texttt{cv()} in this case are the same as for a model fit by
\texttt{lmer()} or \texttt{glmer()}. We illustrate with the mixed model
fit to the HSB data:

@@

We used the same random-number generator seeds as in the previous
example cross-validating the model fit by \texttt{lmer()}, and so the
same folds are employed in both cases.\footnote{The observant reader
  will notice that we set the argument
  \texttt{control=list(opt="optim")} in the call to \texttt{lme()},
  changing the optimizer employed from the default \texttt{"nlminb"}. We
  did this because with the default optimizer, \texttt{lme()}
  encountered the same convergence issue as \texttt{lmer()}, but rather
  than issuing a warning, \texttt{lme()} failed, reporting an error. As
  it turns out, setting the optimizer to \texttt{"optim"} avoids this
  problem.} The estimated covariance components and fixed effects in the
summary output differ slightly between the \texttt{lmer()} and
\texttt{lme()} solutions, although both functions seek to maximize the
REML criterion. This is, of course, to be expected when different
algorithms are used for numerical optimization. To the precision
reported, the cluster-level CV results for the \texttt{lmer()} and
\texttt{lme()} models are identical, while the case-level CV results are
very similar but not identical.

@@

We start by generating the data, using parameters consistent with the
description above and meant to highlight the issues that arise in
cross-validating mixed-effects models:\footnote{The
  \texttt{regsubsets()} function computes several measures of model
  predictive performance, including the \(R^2\) and \(R^2\) adjusted for
  degrees of freedom, the residual sums of squares, Mallows's \(C_p\),
  and the BIC. Several of these are suitable for comparing models with
  differing numbers of coefficients---we use the BIC below---but all
  necessarily agree when comparing models with the \emph{same} number of
  coefficients.}

@@

Here is a scatterplot of the data for a representative group of 10
(without loss of generality, the first 10) of 100 students, showing the
95\% concentration ellipse for each cluster:\footnote{We find it
  convenient to use the \textbf{lattice} \citep{Sarkar:2008} and
  \textbf{latticeExtra} \citep{SarkarAndrews:2022} packages for this and
  other graphs in this section.}

@@

We proceed to obtain predictions from the random-intercept model
(\texttt{mod.0}) and the other models (\texttt{mod.1}, \texttt{mod.2},
and \texttt{mod.3}) based on fixed effects alone, as would be used for
cross-validation based on clusters (i.e., students), and for fixed and
random effects---so-called best linear unbiased predictions or
BLUPs---as would be used for cross-validation based on cases (i.e.,
occasions within students):

@@

Predictions based on the random-intercept model \texttt{mod.0} for the
first 10 students are shown in the following graph:

@@

Our next model, \texttt{mod.1}, includes a fixed intercept and fixed
effect of \texttt{x} along with a random intercept:

@@

The BLUPs fit the observed data very closely, but predictions based on
the fixed effects alone, with a common intercept and slope for all
clusters, are very poor---indeed, much worse than the fixed-effects-only
predictions based on the simpler random-intercept model, \texttt{mod.0}.
We therefore anticipate (and show later in this section) that case-based
cross-validation will prefer \texttt{mod1} to \texttt{mod0}, but that
cluster-based cross-validation will prefer \texttt{mod0} to
\texttt{mod1}.

Our third model, \texttt{mod.2}, includes the contextual effect of
\(x\)---that is, the cluster mean \texttt{xm}---along with \(x\) and the
intercept in the fixed-effect part of the model, and a random intercept:

@@

This model is equivalent to fitting
\texttt{y\ \textasciitilde{}\ I(x\ -\ xm)\ +\ xm\ +\ (1\ \textbar{}\ group)},
which is the model that generated the data once the coefficient of the
contextual predictor \texttt{xm} is set to 0 (as it is in
\texttt{mod.3}, discussed below).

Predictions from model \texttt{mod.2} appear in the following graph:

@@

Depending on the estimated variance parameters of the model, a mixed
model like \texttt{mod.2} will apply varying degrees of shrinkage to the
random-intercept BLUPs that correspond to variation in the heights of
the parallel fitted lines for the individual students. In our contrived
data, the \texttt{mod.2} applies little shrinkage, allowing substantial
variability in the heights of the fitted lines, which closely approach
the observed values for each student. The fit of the mixed model
\texttt{mod.2} is consequently similar to that of a fixed-effects model
with age and a categorical predictor for individual students (i.e.,
treating students as a factor, and not shown here).

The mixed model \texttt{mod.2} therefore fits individual observations
well, and we anticipate a favorable assessment using individual-based
cross-validation. In contrast, the large variability in the BLUPs
results in larger residuals for predictions based on fixed effects
alone, and so we expect that cluster-based cross-validation won't show
an advantage for model \texttt{mod.2} compared to the smaller model
\texttt{mod.0}, which includes only fixed and random intercepts.

@@

Finally, \texttt{mod.3} directly estimates the model used to generate
the data. As mentioned, it is a constrained version of \texttt{mod.2},
with the coefficient of \texttt{xm} set to 0, and with \texttt{x}
expressed as a deviation from the cluster mean \texttt{xm}:

@@

The predictions from \texttt{mod.3} are therefore similar to those from
\texttt{mod.2}:

@@

We next carry out case-based cross-validation, which, as we have
explained, is based on both fixed and predicted random effects (i.e.,
BLUPs), and cluster-based cross-validation, which is based on fixed
effects only. In order to reduce between-model random variability in
comparisons of models, we apply \texttt{cv()} to the list of models
created by the \texttt{models()} function (introduced previously),
performing cross-validation with the same folds for each model:

@@

We'll illustrate with data on pig growth, introduced by \citet[Table
3.1]{DiggleLiangZeger:1994}. The data are in the \texttt{Pigs} data
frame in the \textbf{cv} package:

@@

The other arguments to \texttt{cvSelect()} are:

>>>>

\author{
John Fox~\orcidlink{0000-0002-1196-8012}\\McMaster
University \And Georges Monette~\orcidlink{0000-0000-0000-0000}\\York
University
}
\title{\pkg{cv}: An \proglang{R} Package for Cross-Validation of
Regression Models}

@@

The data for this example are drawn from the \pkg{ISLR2} package for
\proglang{R}, associated with \citet{JamesEtAl:2021}. The presentation
here is close (though not identical) to that in the original source
\citep[ Secs. 5.1, 5.3]{JamesEtAl:2021}, and it demonstrates the use of
the \texttt{cv()} function in the \pkg{cv} package.\footnote{\citet{JamesEtAl:2021}
  use the \texttt{cv.glm()} function in the \pkg{boot} package
  \citep{CantyRipley2022, DavisonHinkley:1997}. Despite its name,
  \texttt{cv.glm()} is an independent function and not a method of a
  \texttt{cv()} generic function.}

@@

The code for this graph uses the \texttt{mse()} function from the
\pkg{cv} package to compute the MSE for each fit.

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 44768
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.52622
bias-adjusted cross-validation criterion = 19.49747
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

The \code{"lm"} method by default uses \code{mse()} as the CV criterion
and the Woodbury matrix identity to update the regression with each fold
deleted without having literally to refit the model. Computational
details are discussed in the final section of this vignette. The
function reports the CV estimate of MSE, a biased-adjusted estimate of
the MSE (the bias adjustment is explained in the final section), and the
MSE is also computed for the original, full-sample regression. Because
the division of the data into 10 folds is random, \code{cv()} explicitly
(randomly) generates and saves a seed for \proglang{R}'s pseudo-random
number generator, to make the results replicable. The user can also
specify the seed directly via the \code{seed} argument to \code{cv()}.

To perform LOO CV, we can set the \code{k} argument to \code{cv()} to
the number of cases in the data, here \code{k=392}, or, more
conveniently, to \code{k="loo"} or \code{k="n"}:

@@

For LOO CV of a linear model, \code{cv()} by default uses the hatvalues
from the model fit to the full data for the LOO updates, and reports
only the CV estimate of MSE. Alternative methods are to use the Woodbury
matrix identity or the ``naive'' approach of literally refitting the
model with each case omitted. All three methods produce exact results
for a linear model (within the precision of floating-point
computations):

@@

The \code{"naive"} and \code{"Woodbury"} methods also return the
bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an
issue for LOO CV.

This is a small regression problem and all three computational
approaches are essentially instantaneous, but it is still of interest to
investigate their relative speed. In this comparison, we include the
\code{cv.glm()} function from the \pkg{boot} package, which takes the
naive approach, and for which we have to fit the linear model as an
equivalent Gaussian GLM. We use the \code{microbenchmark()} function
from the package of the same name for the timings \citep{Mersmann:2023}:

@@

On our computer, using the hatvalues is about an order of magnitude
faster than employing Woodbury matrix updates, and more than two orders
of magnitude faster than refitting the model.\footnote{Out of
  impatience, we asked \code{microbenchmark()} to execute each command
  only 10 times rather than the default 100. With the exception of the
  last columns, the output is self-explanatory. The last column shows
  which methods have average timings that are statistically
  distinguishable. Because of the small number of repetitions (i.e.,
  10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't
  distinguishable, but the difference between these methods persists
  when we perform more repetitions---we invite the reader to redo this
  computation with the default \code{times=100} repetitions.}

@@

The \code{cv()} function also has a method that can be applied to a list
of regression models for the same data, composed using the
\code{models()} function. For \(k\)-fold CV, the same folds are used for
the competing models, which reduces random error in their comparison.
This result can also be obtained by specifying a common seed for
\proglang{R}'s random-number generator while applying \code{cv()}
separately to each model, but employing a list of models is more
convenient for both \(k\)-fold and LOO CV (where there is no random
component to the composition of the \(n\) folds).

We illustrate with the polynomial regression models of varying degree
for the \code{Auto} data (discussed previously), beginning by fitting
and saving the 10 models:

@@

We then apply \code{cv()} to the list of 10 models (the \code{data}
argument is required):

@@

Because we didn't supply names for the models in the calls to the
\code{models()} function, the names \code{model.1}, \code{model.2},
etc., are generated by the function.

@@

Alternatively, we can use the \code{plot()} method for
\code{"cvModList"} objects to compare the models, though with separate
graphs for 10-fold and LOO CV:

@@

The \code{Mroz} data set from the \pkg{carData} package
\citep[associated with][]{FoxWeisberg:2019} has been used by several
authors to illustrate binary logistic regression; see, in particular
\citet{FoxWeisberg:2019}. The data were originally drawn from the U.S.
Panel Study of Income Dynamics and pertain to married women. Here are a
few cases in the data set:

@@

The response variable in the logistic regression is \code{lfp},
labor-force participation, a factor coded \code{"yes"} or \code{"no"}.
The remaining variables are predictors:

\begin{itemize}
\tightlist
\item
  \code{k5}, number of children 5 years old of younger in the woman's
  household;
\item
  \code{k618}, number of children between 6 and 18 years old;
\item
  \code{age}, in years;
\item
  \code{wc}, wife's college attendance, \code{"yes"} or \code{"no"};
\item
  \code{hc}, husband's college attendance;
\item
  \code{lwg}, the woman's log wage rate if she is employed, or her
  \emph{imputed} wage rate, if she is not \citep[a variable that][ show
  is problematically defined]{FoxWeisberg:2019}; and
\item
  \code{inc}, family income, in \$1000s, exclusive of wife's income.
\end{itemize}

We use the \code{glm()} function to fit a binary logistic regression to
the \code{Mroz} data:

@@

In addition to the usually summary output for a GLM, we show the result
of applying the \code{BayesRule()} function from the \pkg{cv} package to
predictions derived from the fitted model. Bayes rule, which predicts a
``success'' in a binary regression model when the fitted probability of
success {[}i.e., \(\phi = \Pr(y = 1)\){]} is \(\widehat{\phi} \ge .5\)
and a ``failure'' if \(\widehat{\phi} < .5\).\footnote{\code{BayesRule()}
  does some error checking; \code{BayesRule2()} is similar, but omits
  the error checking, and so can be faster for large problems.} The
first argument to \code{BayesRule()} is the binary \{0, 1\} response,
and the second argument is the predicted probability of success.
\code{BayesRule()} returns the proportion of predictions that are
\emph{in error}, as appropriate for a ``cost'' function.

The value returned by \code{BayesRule()} is associated with an
``attribute'' named \code{"casewise loss"} and set to
\code{"y != round(yhat)"}, signifying that the Bayes rule CV criterion
is computed as the mean of casewise values, here 0 if the prediction for
a case matches the observed value and 1 if it does not (signifying a
prediction error). The \code{mse()} function for numeric responses is
also calculated as a casewise average. Some other criteria, such as the
median absolute error, computed by the \code{medAbsErr()} function in
the \pkg{cv} package, aren't averages of casewise components. The
distinction is important because, to our knowledge, the statistical
theory of cross-validation, for example, in \citet{DavisonHinkley:1997},
\citet{BatesHastieTibshirani:2023}, and \citet{ArlotCelisse:2010}, is
developed for CV criteria like MSE that are means of casewise
components. As a consequence, we limit computation of bias adjustment
and confidence intervals (see below) to criteria that are casewise
averages.

@@

The \code{"glm"} method for \code{cv()} is largely similar to the
\code{"lm"} method, although the default algorithm, selected explicitly
by \code{method="exact"}, refits the model with each fold removed (and
is thus equivalent to \code{method="naive"} for \code{"lm"} models). For
generalized linear models, \code{method="Woodbury"} or (for LOO CV)
\code{method="hatvalues"} provide approximate results (see the last
section of the vignette for details):

@@

To ensure that the two methods use the same 10 folds, we specify the
seed for \proglang{R}'s random-number generator explicitly; here, and as
is common in our experience, the \code{"exact"} and \code{"Woodbury"}
algorithms produce nearly identical results. The CV estimates of
prediction error are slightly higher than the estimate based on all of
the cases.

The printed output includes a 95\% confidence interval for the
bias-adjusted Bayes rule CV criterion.
\citet{BatesHastieTibshirani:2023} show that these confidence intervals
are unreliable for models fit to small samples, and by default
\code{cv()} computes them only when the sample size is 400 or larger and
when the CV criterion employed is an average of casewise components, as
is the case for Bayes rule. See the final section of the vignette for
details of the computation of confidence intervals for bias-adjusted CV
criteria.

@@

As for linear models, we report some timings for the various
\textbackslash code\{cv()` methods of computation in LOO CV as well as
for the \code{cv.glm()} function from the \pkg{boot} package (which,
recall, refits the model with each case removed, and thus is comparable
to \code{cv()} with \code{method="exact"}):

@@

In contrast, mixed-effects models are fit to \emph{dependent} data, in
which cases as clustered, such as hierarchical data, where the clusters
comprise higher-level units (e.g., students clustered in schools), or
longitudinal data, where the clusters are individuals and the cases
repeated observations on the individuals over time.\footnote{There are,
  however, more complex situations that give rise to so-called
  \emph{crossed} (rather than \emph{nested}) random effects. For
  example, consider students within classes within schools. In primary
  schools, students typically are in a single class, and so classes are
  nested within schools. In secondary schools, however, students
  typically take several classes and students who are together in a
  particular class may not be together in other classes; consequently,
  random effects based on classes within schools are crossed. The
  \code{lmer()} function in the \pkg{lme4} package is capable of
  modeling both nested and crossed random effects, and the \code{cv()}
  methods for mixed models in the \pkg{cv} package pertain to both
  nested and crossed random effects. We present an example of the latter
  later in the vignette.}

@@

The HSB data are included in the \code{MathAchieve} and
\code{MathAchSchool} data sets in the \pkg{nlme} package
\citep{PinheiroBates:2000}. \code{MathAchieve} includes individual-level
data on 7185 students in 160 high schools, and \code{MathAchSchool}
includes school-level data:

@@

We'll use only the \code{School}, \code{SES} (students' socioeconomic
status), and \code{MathAch} (their score on a standardized
math-achievement test) variables in the \code{MathAchieve} data set, and
\code{Sector} (\code{"Catholic"} or \code{"Public"}) in the
\code{MathAchSchool} data set.

Some data-management is required before fitting a mixed-effects model to
the HSB data, for which we use the \pkg{dplyr} package
\citep{WickhamEtAl:2023}:

@@

In the process, we created two new school-level variables:
\code{meanses}, which is the average SES for students in each school;
and \code{cses}, which is school-average SES centered at its mean. For
details, see \citet[Sec. 7.2.2]{FoxWeisberg:2019}.

Still following Fox and Weisberg, we proceed to use the \code{lmer()}
function in the \pkg{lme4} package \citep{BatesEtAl:2015} to fit a mixed
model for math achievement to the HSB data:

@@

For cluster-level CV, the \code{clusterVariables} argument tells
\code{cv()} how the clusters are defined. Were there more than one
clustering variable, say classes within schools, these would be provided
as a character vector of variable names:
\code{clusterVariables = c("school", "class")}. For cluster-level CV,
the default is \code{k = "loo"}, that is, leave one cluster out at a
time; we instead specify \code{k = 10} folds of clusters, each fold
therefore comprising \(160/10 = 16\) schools.

If the \code{clusterVariables} argument is omitted, then case-level CV
is employed, with \code{k = 10} folds as the default, here each with
\(7185/10 \approx 719\) students. Notice that one of the 10 models refit
with a fold removed failed to converge. Convergence problems are common
in mixed-effects modeling. The apparent issue here is that an estimated
variance component is close to or equal to 0, which is at a boundary of
the parameter space. That shouldn't disqualify the fitted model for the
kind of prediction required for cross-validation.

There is also a \code{cv()} method for linear mixed models fit by the
\code{lme()} function in the \pkg{nlme} package, and the arguments for
\code{cv()} in this case are the same as for a model fit by
\code{lmer()} or \code{glmer()}. We illustrate with the mixed model fit
to the HSB data:

@@

We used the same random-number generator seeds as in the previous
example cross-validating the model fit by \code{lmer()}, and so the same
folds are employed in both cases.\footnote{The observant reader will
  notice that we set the argument \code{control=list(opt="optim")} in
  the call to \code{lme()}, changing the optimizer employed from the
  default \code{"nlminb"}. We did this because with the default
  optimizer, \code{lme()} encountered the same convergence issue as
  \code{lmer()}, but rather than issuing a warning, \code{lme()} failed,
  reporting an error. As it turns out, setting the optimizer to
  \code{"optim"} avoids this problem.} The estimated covariance
components and fixed effects in the summary output differ slightly
between the \code{lmer()} and \code{lme()} solutions, although both
functions seek to maximize the REML criterion. This is, of course, to be
expected when different algorithms are used for numerical optimization.
To the precision reported, the cluster-level CV results for the
\code{lmer()} and \code{lme()} models are identical, while the
case-level CV results are very similar but not identical.

@@

We start by generating the data, using parameters consistent with the
description above and meant to highlight the issues that arise in
cross-validating mixed-effects models:\footnote{The \code{regsubsets()}
  function computes several measures of model predictive performance,
  including the \(R^2\) and \(R^2\) adjusted for degrees of freedom, the
  residual sums of squares, Mallows's \(C_p\), and the BIC. Several of
  these are suitable for comparing models with differing numbers of
  coefficients---we use the BIC below---but all necessarily agree when
  comparing models with the \emph{same} number of coefficients.}

@@

Here is a scatterplot of the data for a representative group of 10
(without loss of generality, the first 10) of 100 students, showing the
95\% concentration ellipse for each cluster:\footnote{We find it
  convenient to use the \pkg{lattice} \citep{Sarkar:2008} and
  \pkg{latticeExtra} \citep{SarkarAndrews:2022} packages for this and
  other graphs in this section.}

@@

We proceed to obtain predictions from the random-intercept model
(\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and
\code{mod.3}) based on fixed effects alone, as would be used for
cross-validation based on clusters (i.e., students), and for fixed and
random effects---so-called best linear unbiased predictions or
BLUPs---as would be used for cross-validation based on cases (i.e.,
occasions within students):

@@

Predictions based on the random-intercept model \code{mod.0} for the
first 10 students are shown in the following graph:

@@

Our next model, \code{mod.1}, includes a fixed intercept and fixed
effect of \code{x} along with a random intercept:

@@

The BLUPs fit the observed data very closely, but predictions based on
the fixed effects alone, with a common intercept and slope for all
clusters, are very poor---indeed, much worse than the fixed-effects-only
predictions based on the simpler random-intercept model, \code{mod.0}.
We therefore anticipate (and show later in this section) that case-based
cross-validation will prefer \code{mod1} to \code{mod0}, but that
cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.

Our third model, \code{mod.2}, includes the contextual effect of
\(x\)---that is, the cluster mean \textbackslash code\{xm{]}---along
with \(x\) and the intercept in the fixed-effect part of the model, and
a random intercept:

@@

This model is equivalent to fitting
\code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that
generated the data once the coefficient of the contextual predictor
\code{xm} is set to 0 (as it is in \textbackslash code\{mod.3{]},
discussed below).

Predictions from model \code{mod.2} appear in the following graph:

@@

Depending on the estimated variance parameters of the model, a mixed
model like \code{mod.2} will apply varying degrees of shrinkage to the
random-intercept BLUPs that correspond to variation in the heights of
the parallel fitted lines for the individual students. In our contrived
data, the \code{mod.2} applies little shrinkage, allowing substantial
variability in the heights of the fitted lines, which closely approach
the observed values for each student. The fit of the mixed model
\code{mod.2} is consequently similar to that of a fixed-effects model
with age and a categorical predictor for individual students (i.e.,
treating students as a factor, and not shown here).

The mixed model \code{mod.2} therefore fits individual observations
well, and we anticipate a favorable assessment using individual-based
cross-validation. In contrast, the large variability in the BLUPs
results in larger residuals for predictions based on fixed effects
alone, and so we expect that cluster-based cross-validation won't show
an advantage for model \code{mod.2} compared to the smaller model
\code{mod.0}, which includes only fixed and random intercepts.

@@

Finally, \code{mod.3} directly estimates the model used to generate the
data. As mentioned, it is a constrained version of \code{mod.2}, with
the coefficient of \code{xm} set to 0, and with \code{x} expressed as a
deviation from the cluster mean \code{xm}:

@@

The predictions from \code{mod.3} are therefore similar to those from
\code{mod.2}:

@@

We next carry out case-based cross-validation, which, as we have
explained, is based on both fixed and predicted random effects (i.e.,
BLUPs), and cluster-based cross-validation, which is based on fixed
effects only. In order to reduce between-model random variability in
comparisons of models, we apply \code{cv()} to the list of models
created by the \code{models()} function (introduced previously),
performing cross-validation with the same folds for each model:

@@

We'll illustrate with data on pig growth, introduced by \citet[Table
3.1]{DiggleLiangZeger:1994}. The data are in the \code{Pigs} data frame
in the \pkg{cv} package:

@@

The other arguments to \code{cvSelect()} are:
